---
title: "Homework 2"
author: "Christina Chen"
date: "2026-10-04"
output: github_document
---

Loading packages
```{r}
library(tidyverse)
```

# Problem 1

Reading in the csv data file, cleaning names, keeping the variables we want, and converting the entry variable from character to logical.
```{r}
transit_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("","NA",".")) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = recode_values(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    ))
```

This dataset currently includes the following variables: line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entrance_type, entry, vending, ada. These variables contain values for the station line, its name, location, routes that it serves, entrance type, entry, vending, and ADA compliance. I read in the original csv file and cleaned the variable names so they were in snake case. Then, since it originally contained extra information that I did not need (such as the division, staff and staff hours, and exit type), I just kept the variables listed above. Lastly, I changed the entry variable from a character (YES/NO) to logical (TRUE/FALSE).

The dataset is not tidy, since the routes served are separate columns route1-11, rather than under one column. This means that those columns are values, rather than variable names.

```{r}
transit_df |> 
  distinct(station_name,line) |> 
  nrow()
```
There are 465 distinct stations.

```{r}
transit_df |> 
  filter(ada == TRUE) |> 
  distinct(station_name,line) |> 
  nrow()
```
There are 84 distinct stations that are ADA compliant.

Total number of station entrances/exits without vending:
```{r}
transit_df |> 
  filter(vending == "NO") |> 
  nrow()
```
Number of station entrances/exits without vending and allow entry:
```{r}
transit_df |> 
  filter(entry == TRUE, vending == "NO") |> 
  nrow()
```

69/183 of the station entrances/exits without vending allow entrance. 






