---
title: "Homework 2"
author: "Christina Chen"
date: "2026-10-04"
output: github_document
---

Loading packages
```{r message = F}
library(tidyverse)
```

# Problem 1

### Reading in the csv data file, cleaning names, keeping the variables we want, and converting the entry variable from character to logical:
```{r}
transit_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("","NA",".")) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = recode_values(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    ))
```

This dataset currently includes the following variables: line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entrance_type, entry, vending, ada. These variables contain values for the station line, its name, location, routes that it serves, entrance type, entry, vending, and ADA compliance. I read in the original csv file and cleaned the variable names so they were in snake case. Then, since it originally contained extra information that I did not need (such as the division, staff and staff hours, and exit type), I just kept the variables listed above. Lastly, I changed the entry variable from a character (YES/NO) to logical (TRUE/FALSE).

The dataset is not tidy, since the routes served are separate columns route1-11, rather than under one column. This means that those columns are values, rather than variable names.

### Answering some questions about the data:
*How many distinct stations are there?*
```{r}
transit_df |> 
  distinct(station_name,line) |> 
  nrow()
```
There are 465 distinct stations.

*How many stations are ADA compliant?*
```{r}
transit_df |> 
  filter(ada == TRUE) |> 
  distinct(station_name,line) |> 
  nrow()
```
There are 84 distinct stations that are ADA compliant.


*What proportion of station entrances / exits without vending allow entrance?*

Total number of station entrances/exits without vending:
```{r}
transit_df |> 
  filter(vending == "NO") |> 
  nrow()
```
Number of station entrances/exits without vending and allow entry:
```{r}
transit_df |> 
  filter(entry == TRUE, vending == "NO") |> 
  nrow()
```
69/183 of the station entrances/exits without vending allow entrance. 

### Reformat the data so that route number and route name are distinct variables, then answer some questions:
```{r}
transit_df_tidy =
  transit_df |> 
  mutate(route8 = as.character(route8),
         route9 = as.character(route9),
         route10 = as.character(route10),
         route11 = as.character(route11)
           ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
    )
```

*How many distinct stations serve the A train?*
```{r}
transit_df_tidy |> 
  filter(route_name == "A") |> 
  distinct(station_name,line) |> 
  nrow()
```
60 distinct stations serve the A train.

*Of the stations that serve the A train, how many are ADA compliant?*
```{r}
transit_df_tidy |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(station_name,line) |> 
  nrow()
```
Of the stations that serve the A train, 17 of them are ADA compliant.






