---
title: "Homework 2"
author: "Christina Chen"
date: "2026-10-04"
output: github_document
---

Loading packages:
```{r message = F}
library(tidyverse)
library(readxl)
```

# Problem 1

**Reading in the csv data file, cleaning names, keeping the variables we want, and converting the entry variable from character to logical:**
```{r message = F}
transit_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("","NA",".")) |> 
  janitor::clean_names() |> 
  select(line:entry, vending, ada) |> 
  mutate(
    entry = recode_values(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    ))
```

This dataset currently includes the following variables: line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entrance_type, entry, vending, ada. These variables contain values for the station line, its name, location, routes that it serves, entrance type, entry, vending, and ADA compliance. I read in the original csv file and cleaned the variable names so they were in snake case. Then, since it originally contained extra information that I did not need (such as the division, staff and staff hours, and exit type), I just kept the variables listed above. Lastly, I changed the entry variable from a character (YES/NO) to logical (TRUE/FALSE).

The dataset is not tidy, since the routes served are separate columns route1-11, rather than under one column. This means that those columns are values, rather than variable names.

**Answering some questions about the data:**
*How many distinct stations are there?*
```{r}
transit_df |> 
  distinct(station_name,line) |> 
  nrow()
```
There are 465 distinct stations.

*How many stations are ADA compliant?*
```{r}
transit_df |> 
  filter(ada == TRUE) |> 
  distinct(station_name,line) |> 
  nrow()
```
There are 84 distinct stations that are ADA compliant.


*What proportion of station entrances / exits without vending allow entrance?*

Total number of station entrances/exits without vending:
```{r}
no_vending =
  transit_df |> 
    filter(vending == "NO") |> 
    nrow()
```
Number of station entrances/exits without vending and allow entry:
```{r}
allow_entry =
  transit_df |> 
    filter(entry == TRUE, vending == "NO") |> 
    nrow()
```
`r allow_entry`/`r no_vending`, or `r round(allow_entry/no_vending, digits = 4)` of the station entrances/exits without vending allow entrance.


**Reformat the data so that route number and route name are distinct variables, then answer some questions:**
```{r}
transit_df_tidy =
  transit_df |> 
  mutate(route8 = as.character(route8),
         route9 = as.character(route9),
         route10 = as.character(route10),
         route11 = as.character(route11)
           ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
    )
```

*How many distinct stations serve the A train?*
```{r}
transit_df_tidy |> 
  filter(route_name == "A") |> 
  distinct(station_name,line) |> 
  nrow()
```
60 distinct stations serve the A train.

*Of the stations that serve the A train, how many are ADA compliant?*
```{r}
transit_df_tidy |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(station_name,line) |> 
  nrow()
```
Of the stations that serve the A train, 17 of them are ADA compliant.

# Problem 2

Read and clean the Mr. Trash Wheel dataset:
```{r}
mr_df =
  read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N742") |> 
  janitor::clean_names() |> 
  mutate(sports_balls = as.integer(sports_balls),
         plastic_bottles = as.character(plastic_bottles))
```

Read and clean Professor Trash Wheel and Gwynnda datasets and add wheel variable to mark which wheel the data belong to
(Note: the variable `plastic_bottles` had to be converted to character to make sure they were aligned across all wheels before binding):
```{r}
prof_df = 
  read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M139") |> 
  janitor::clean_names() |> 
  mutate(wheel = "professor",
         year = as.character(year))

gwynn_df =
read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda the Good Wheel of the W", range = "A2:L394") |>
  janitor::clean_names() |> 
  mutate(wheel = "gwynnda",
         year = as.character(year),
         plastic_bottles = as.character(plastic_bottles))

mr_df = mutate(mr_df, wheel = "mr")
```

Bind all the datasets together:
```{r}
trash_wheel_tidy =
  bind_rows(mr_df, prof_df, gwynn_df) |> 
  relocate(wheel)
```


The tidied dataset `trash_wheel_tidy` includes information on `r nrow(distinct(trash_wheel_tidy, wheel))` trash wheels: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda the Good Wheel of the West. There are `r nrow(trash_wheel_tidy)` observations in this dataset and `r ncol(trash_wheel_tidy)` variables. Some key variables in this dataset include the date of observation, weight of trash in tons, volume of trash in cubic yards, number of different types of trash collected, and number of Maryland homes powered by the trash. The total weight of trash collected by Professor Trash Wheel is `r trash_wheel_tidy |> filter(wheel == "professor") |> pull(weight_tons) |> sum()` tons. The total number of cigarette butts collected by Gwynnda in June of 2022 is `r trash_wheel_tidy |> filter(wheel == "gwynnda", month == "June", year == "2022") |> pull(cigarette_butts) |> sum()` cigarette butts. 

# Problem 3

Initial import, cleaning, and tidying of baseline dataset:
```{r message = F}
baseline_df =
  read_csv("data/data_mci/MCI_baseline.csv", skip = 1, na = c("","NA",".")) |> 
  janitor::clean_names() |> 
  mutate(sex = recode(sex, `1` = "male", `0` = "female"),
         apoe4 = recode(apoe4, `1` = "carrier", `0` = "non-carrier")) |> 
  filter(is.na(age_at_onset) | age_at_onset >= current_age)
```

* Import baseline dataset, which is a csv file, skipping the first row, which is a header.
* Clean variable names using janitor package
* Recode `sex` and `apoe4` so they are not numeric and instead have labeled categories
* Remove participants that have MCI at baseline.

The baseline dataset `baseline_df` provides information on the participants' id, baseline age, sex, years of education, whether they are a carrier of APOE4, and age of onset of MCI if they did get MCI during the follow-up. This dataset has `r nrow(baseline_df)` observations or participants recruited, and `r ncol(baseline_df)` variables. Of these participants, `r nrow(filter(baseline_df, age_at_onset != "na"))` participants develop MCI. The average baseline age is `r round(mean(baseline_df$current_age), digits = 2)` years. `r nrow(filter(baseline_df, sex == "female", apoe4 == "carrier"))`/`r nrow(filter(baseline_df, sex == "female"))`, or `r nrow(filter(baseline_df, sex == "female", apoe4 == "carrier"))/nrow(filter(baseline_df, sex == "female"))` of the women in the study are APOE4 carriers.

Import, clean, tidying of longitudinal biomarker dataset:
```{r message = F}
long_bio_df =
  read_csv("data/data_mci/mci_amyloid.csv", skip = 1, na = c("","NA",".")) |> 
  janitor::clean_names() |> 
  rename(id = study_id)
```

* Import longitudinal biomarker dataset, which is a csv file
* Clean variable names using janitor package
* Rename `study_id` column to `id` to match baseline dataset

The longitudinal biomarker dataset `long_bio_df` provides information on amyloid beta 42/40 ratio (an Alzheimer's biomarker) at baseline and timepoints 2, 4, 6, and 8 of the study for each participant. This dataset has `r nrow(long_bio_df)` observations or participants in this dataset, and `r ncol(long_bio_df)` variables.

Some participants do appear in only the baseline or amyloid datasets. 

Combine baseline and longitudinal biomarker datasets so that only participants in both datasets are there:
```{r}
combined_df =
  inner_join(baseline_df, long_bio_df, by = "id")
```

After combining `baseline_df` and `long_bio_df` using `inner_join` to keep only participants in both datasets, we have a combined dataset with `r nrow(combined_df)` observations or participants in this dataset, and `r ncol(combined_df)` variables.

Exporting resulting dataframe:
```{r}
write_csv(combined_df, "data/data_mci/mci_combined.csv")
```











